plyranges: A grammar of genomic data transformation

### 2018-09-20 paper.Rmd
code comparison and latex fixes

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \\textit{x} and \\textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \\textit{x} that have been overlapped 
(the rectangles of \\textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \\textit{y} that overlap 
the ranges in \\textit{x}. In this case a range in \\textit{y} overlaps the 
ranges in \\textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by
\\textit{y} and merge all metadata columns from \\textit{x} and \\textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour
operations. The join semantics alter what is returned: \\textbf{A}: for an 
\\textbf{inner} join the \\textit{x} ranges that are overlapped by \\textit{y} 
are returned. The returned ranges also include the metadata from the \\textit{y}
range that overlapped the three \\textit{x} ranges. \\textbf{B} An 
\\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \\textit{x} ranges and the 
\\textit{y} ranges. \\textbf{C} For the left join \\textbf{left} join all 
\\textit{x} ranges are returned regardless of whether they are overlapped by 
\\textit{y}. In this case the resulting third range  of the join would have 
missing values on metadata columns that came from \\textit{y}.}
\label{fig:olaps-fig}
\end{figure}

@@

@@

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar. Consequently, `plyranges` code is generally more compact
than the equivalent `GenomicRanges` code (figure \ref{code-fig}).

>>>>

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \textit{x} that have been overlapped 
(the rectangles of \textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \textit{y} that overlap 
the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
ranges in \textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \texttt{plyranges} DSL for both overlap and nearest neighbour
operations. The join semantics alter what is returned: \textbf{A}: for an 
\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
are returned. The returned ranges also include the metadata from the \textit{y}
range that overlapped the three \textit{x} ranges. \textbf{B} An 
\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \textit{x} ranges and the 
\textit{y} ranges. \\textbf{C} For the left join \textbf{left} join all 
\textit{x} ranges are returned regardless of whether they are overlapped by 
\textit{y}. In this case the resulting third range  of the join would have 
missing values on metadata columns that came from \textit{y}.}
\label{fig:olaps-fig}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
\caption{Idiomatic code examples for \texttt{plyranges} (\textbf(A)) and 
\texttt{GenomicRanges} (\textbf(B)) illustrating an overlap and aggregate 
operation that returns the same result. 
In each example, we have two BED files consisting of SNPs that are
genome-wide association study (GWAS) hits and reference exons. Each code block
counts for each SNP the number of distinct exons it overlaps. The 
\texttt(plyranges) code achieves this with an overlap join followed by partitioning
and aggregation. Strand is ignored by default here. 
The \texttt(GenomicRanges) code achieves this using the 'Hits' and 'List' 
classes and their methods.}
\label{fig:code}
\end{figure}

@@

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar. Consequently, `plyranges` code is generally more compact
than the equivalent `GenomicRanges` code (figure \ref{fig:code}).
